Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models

### 2018-01-08 paper-amia-summit-2018/amia_summit_2018.tex
update paper with reviewrs comments

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics. In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges. Our method achieved 70.38\%, 60.67\%, 86.77\% F1-score for under-sampled sequences and 77.75\%, 75.20\%, 83.81\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively. The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. Sequence classification methods are utilized for classifying temporally ordered sequences. Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}. In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease. Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.

%Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection. In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information is learned from the dataset and used as a ``model'' to classify sequences.

%This makes sequence classification a more challenging task than traditional classification.

One of the model-based classification methods is a probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM). Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory. It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}. A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}. In Lipton et. al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc. LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit. A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed. These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.

Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

%This work is the next step in a series of experiments to develop the [informatics based methods described above]. 
In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
We propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into client codes (adolescent and caregiver) group and counselor codes group. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Here, the successful sequences had an average length of 9.79 while that of unsuccessful sequences was 9.65. For each of the probabilistic models (Markov chain and HMM), two models were trained, one model was trained using successful sequences and another one was trained using unsuccessful sequences.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. That RNN can capture long-term dependencies for predicting the future events is its main advantage over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
As these equations, $\sigma$ stands for an element-wise application of the sigmoid function and $\odot$ is an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W$ and $U$ represent weights and $b$ represents biases. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn_small.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence. Then, we applied a softmax layer at each time-step for predicting the target which is the predicted label of the sequence. This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences. Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\ 
\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under and over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Undersampling}} & \multicolumn{3}{|c|}{\textbf{Oversampling}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7060 & 0.7044 & 0.7038 & 0.7932 & 0.7799 & 0.7775 \\ \hline
 Markov Chain $2^{nd}$ Order & 0.6395 & 0.6385 & 0.6379 & 0.7111 & 0.7029 & 0.7000\\ \hline
 Hidden Markov Model & 0.6244 & 0.6143 & 0.6067 & 0.7775 & 0.7567 & 0.7520\\ \hline
 LSTM RNN & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
 LSTM RNN - TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
 GRU RNN & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
 GRU RNN - TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
  \end{tabular}
\end{table}  
\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform over probabilistic models and achieved 16.39\%-26.1\% higher F1-score. Second, LSTM with target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower precision, recall, and F-measure than first-order MC. In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first to second-order models.

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset. For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size 55. Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%, 2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain. Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with target replication has the highest F1-score among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences although the performance of probabilistic models increases. We also observed that all models have the largest value in precision compared to other performance metrics.

\begin{table}[h]
\centering
\caption{Most likely communication sequences in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Most likely communication sequences} \\ \hline      
successful & GINFO+: General information, positive $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ OQTBN: \\ 
& Open-ended question, target behavior neutral \\\hline
successful & SS: Structure session $\rightarrow $ GINFO+: General information, positive $\rightarrow $ CQECHT+: Closed-ended \\
& question, elicit change talk positive\\\hline
successful & SO: Statement, other $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ AF: Affirm $\rightarrow $ HUPW: High uptake, \\
& weight $\rightarrow $ OQECML+: Open-ended question, elicit commitment language positive. \\\hline
unsuccessful & ADV+: Advise, positive $\rightarrow $ AMB-: Ambivalence negative $\rightarrow $ OQECHT-: Open-ended \\
& question, elicit change talk negative\\\hline
unsuccessful & CQECHT+: Open-ended question, elicit change talk positive $\rightarrow $ RCHT-S: Reflect, change talk \\
& negative $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative \\\hline
unsuccessful & SUP: Support $\rightarrow $ AF: Affirm $\rightarrow $ CQTBN: Closed-ended question, target behavior neutral \\
& $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative $\rightarrow $ AMB-: Ambivalence negative\\\hline
  \end{tabular}
\end{table}

>>>>

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences generated by a physiological, genomic or psychological process to make predictions about the outcome of that process arises in
many domains of clinical informatics. In this paper, we focus on the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviewing.
Specifically, we formulate the problem of predicting the success of a motivational interview (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of
coded patient-provider communication exchanges as binary sequence classification.
%Severe imbalance between positive and negative training sequences is an additional challenge that arises in the context of motivational interviewing.
We propose two methods that are based on deep learning, such as Recurrent Neural Networks (RNNs), and probabilistic models, such as Markov Chain (MC) and Hidden Markov Model (HMM), and compare the
accuracy of these methods using communication sequences annotated with behavior codes from real-life motivational interviews. Our experiments indicate that the deep learning-based method is
significantly more accurate than the methods based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM,
respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate
that the proposed method can be used for real-time monitoring of the progression of clinical interviews and identification of the most effective communication strategies, which significantly decreases
the effort required to develop psychological interventions and increases their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by genomic, psychological or psychological processes arise in many different domains of clinical informatics (e.g.,
DNA sequences, gene activations, ECG measurements, laboratory results, diagnostic codes, utterances during clinical interviews). Classification (or categorization) is a type of analysis of those
sequences that has a broad range of important practical applications, from protein classification \cite{yakhnenko2005discriminatively} and function prediction \cite{deshpande2002evaluation} to
detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as their temporal order, makes sequence classification a
more challenging task than traditional classification. Predicting the outcome of those sequences (e.g. physiological or behavior response) can also be formulated as sequence classification.

In general, sequence classification methods fall into one of the three major categories: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature
vector and apply a standard supervised learning methods, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
Feature-based methods for sequence classification have had limited success since traditional bag-of-words feature representation methods cannot easily account for the order of observations in a
sequence.
%Shapelet \cite{ye2009time} and pattern based techniques \cite{kudenko1998feature, lesh1999mining} as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of
%standard classifiers as well.
Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is
Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by time points.
The third type of sequence classification methods first creates a probabilistic model, such as Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class
using the training data and then classifies new sequences by applying the constructed models. MC and HMM explicitly capture short-term temporal dependencies (e.g. first- and second-order MC and HMM
\cite{kundu1988recognition}) between adjacent observations in a sequence. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short-
and long-term dependencies and were shown to be effective for modeling different types of sequential data. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which
successfully addresses the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and
handwriting \cite{nion2013handwritten} recognition to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to
recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting
the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
Unit\cite{chung2014empirical} (GRU), was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.

In this paper, we address the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of
Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that
approximately one-third (31.8\%) of US children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and
have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI),
an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires,
ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior
change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of
patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since
manual examination and sequential pattern analysis of transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the
mechanisms of MI's efficacy, in particular.

Our goal is to test the applicability of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for predicting the success of clinical interviews (i.e. eliciting a
particular type of patient behavioral response, such as CHT) and dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on a sequence of
coded previous patient-provider communication exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored
several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook
\cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability
of state-of-the-art methods for sequence modeling to analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavior in a context of motivational
interviews, in particular.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists
of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, in which the codes are grouped into
patient (adolescent and caregiver) codes and counselor codes. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in
positive change talk (CHT+) and commitment language (CML+) statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk (CHT-) or commitment language (CML-) and the
sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Successful sequences had an average length of
9.79 utterances, while unsuccessful sequences had on average 9.65 utterances. For each of the probabilistic models (MC and HMM), two models were trained, one model was trained using successful
sequences and another one was trained using unsuccessful sequences.

@@

\textbf {Recurrent Neural Networks (RNN)} are a type of Neural Networks, which can send feedback from the current hidden state to the hidden state of the next time step. The ability of RNN to capture
long-term dependencies for predicting the future events is its main advantage over MC and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior
observed at some point in the interview is very informative for the future behaviors, that will be observed later, but not immediately. However, it was observed that RNN fails to capture long-term
dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short
Term Memory networks or simply LSTM. There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are
simpler than LSTM units and have been shown to experimentally outperform other models\cite{cho2014properties}. GRU is formally defined as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
As these equations, $\sigma$ stands for an element-wise application of the sigmoid function and $\odot$ is an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W$ and $U$ represent weights and $b$ represents biases. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn_small.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes and embedding of MYSCOPE codes in an observed
sequence of communication behaviors as input. Then, we applied a softmax layer at each time-step for predicting the label of the sequence. Since the sequence label was predicted at each time step, the proposed model is also known as Recurrent Neural Network with Target Replication (RNN-TR). As can be seen from
Eq.~\ref{eq:loss}, the total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where $T$ is the total number of elements in a sequence,
$\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that
best performance achieved when we use $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which
were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges
to optimal after 100 epochs and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences. Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\ 
\begin{table}[ht]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under and over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Undersampling}} & \multicolumn{3}{|c|}{\textbf{Oversampling}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7060 & 0.7044 & 0.7038 & 0.7932 & 0.7799 & 0.7775 \\ \hline
 Markov Chain $2^{nd}$ Order & 0.6395 & 0.6385 & 0.6379 & 0.7111 & 0.7029 & 0.7000\\ \hline
 Hidden Markov Model & 0.6244 & 0.6143 & 0.6067 & 0.7775 & 0.7567 & 0.7520\\ \hline
 LSTM RNN & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
 LSTM RNN - TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
 GRU RNN & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
 GRU RNN - TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
  \end{tabular}
\end{table}  
\subsection*{\textit{Predictive performance for under-sampled Patient-Provider Communication code sequences}}
We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used. Five major conclusions can be drawn
from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform probabilistic models and achieve 16.39\%-26.1\% higher F1-score. Second, LSTM with
target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681. Third, target replication strategy improves the
performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics
for under-sampled sequences. Fifth, second-order MC has lower precision, recall, and F-measure than first-order MC. In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 
9.36\%, when going from first to second-order MC model.

\subsection*{\textit{Predictive performance for over-sampled Patient-Provider Communication code sequences}}
Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset. For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size
55. Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%,
2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain. Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with
target replication has the highest F1-score among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences, while the performance of
probabilistic models increases. We also observed that all models have the largest value in precision compared to other performance metrics.

\begin{table}[ht]
\centering
\caption{Most likely communication sequences in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Most likely communication sequences} \\ \hline      
successful & GINFO+: General information, positive $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ OQTBN: \\ 
& Open-ended question, target behavior neutral \\\hline
successful & SS: Structure session $\rightarrow $ GINFO+: General information, positive $\rightarrow $ CQECHT+: Closed-ended \\
& question, elicit change talk positive\\\hline
successful & SO: Statement, other $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ AF: Affirm $\rightarrow $ HUPW: High uptake, \\
& weight $\rightarrow $ OQECML+: Open-ended question, elicit commitment language positive. \\\hline
unsuccessful & ADV+: Advise, positive $\rightarrow $ AMB-: Ambivalence negative $\rightarrow $ OQECHT-: Open-ended \\
& question, elicit change talk negative\\\hline
unsuccessful & CQECHT+: Open-ended question, elicit change talk positive $\rightarrow $ RCHT-S: Reflect, change talk \\
& negative $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative \\\hline
unsuccessful & SUP: Support $\rightarrow $ AF: Affirm $\rightarrow $ CQTBN: Closed-ended question, target behavior neutral \\
& $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative $\rightarrow $ AMB-: Ambivalence negative\\\hline
  \end{tabular}
\end{table}
